---
title: "Fact Tables"
date: "[Back to home](index.html)"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are **six** fact tables for the Transit Agency Profiles: 

- Ridership (pulled from the DRPT data portal, so no code required here)
  * [DRPT data portal](https://data.drpt.virginia.gov/)
- Service 
- Access to Transit 
  - by Agency 
  - by Route
- Amount of Service
- Census


**A note before we begin: this is a very long document. If you'd like to skip to a certain point in the page, click the table of contents items over to the left**

## Set It Up!

Oops! There's actually some Very Important pre-processing that needs to occur before anything can happen with the analysis. We are going to create a `NovaStops` file with the stop data for all the routes in Northern Va. Using this, we will then create a `NovaStopsRoutes` file that gives us, you guessed it, all the stops and the associated routes. This file also includes one other very important piece of information: the count of trips for each stop. You'll see how we use that a bit later. 



####  Creating `NovaStops`

In case you need to re-run the stop data (perhaps there's been a change in the GTFS? or it's a new year?), you can use the incredibly convenient `stoploc` function from `toolingtransit`. I've included the function here too, in case you can't access `toolingtransit`. 

##### `stoploc`
```{r, eval = F}
stoploc <- function(gtfszip, agency) {
  require(tidyverse)
  require(tidytransit)
  require(sf)
  require(units)
  ### All gtfs data is located in this file path
  GTFS_path <- file.path ("Z:",
                          "NVTC General",
                          "Projects and Programs",
                          "Transit Resource Center (TRC)",
                          "Data",
                          "GTFS")

  #establish gtfs data
  CountyPath <- file.path(GTFS_path, gtfszip)
  CountyGTFS <- read_gtfs(CountyPath)
  #create spatial file for stops and select only necessary variables
  CountyStopLoc <- stops_as_sf(CountyGTFS$stops) %>%
    dplyr::select(stop_id) %>%
    mutate("Agency" = agency)
  return(CountyStopLoc)
}
```

The transit agency profiles include data from ALL our transit agencies. Some of these are easier to come by than others. Let's begin with the easy: 

```{r, eval = F}
#Join 2022 stops
NovaStops_2022 <- rbind(
  stoploc("2022-04_Arlington.zip", "ART"),
  stoploc("2022-03_CUE.zip", "CUE"),
  stoploc("2022-04_DASH.zip", "DASH"),
  stoploc("2022-03_Fairfax_Connector.zip", "FFX"),
  stoploc("2022-07_Loudoun.zip", "LCT"),
  stoploc("2022-03_VRE.zip", "VRE"),
  stoploc("2022-03_OmniRide_PRTC (2).zip", "PRTC")
)

#create columns for Agency name and for a newstop id that includes agency
NovaStops_2022 <- NovaStops_2022 %>%
  unite(stop_id, Agency, col = "newstop_id",remove = F)
```

The only missing transit agency is WMATA. `stoploc` doesn't work here since the Metrobus and Metrorail data need to be clipped to only include Northern Virginia. Metrorail GTFS data also differs from the bus GTFS data, so this was formatted in Excel. In the following chunk, we clip the Metrobus data to Northern Virginia.

You'll see that in the last line of the code, only the new `MetrobusStops` and our previously created `NovaStops_2022` are joined. we will also be adding in `Metrorailstops`, but this will happen a bit later in the process, since it was formatted outside of R.  


```{r, eval = F}
GTFS_path <- file.path ("Z:",
                        "NVTC General",
                        "Projects and Programs",
                        "Transit Resource Center (TRC)",
                        "Data",
                        "GTFS")

Metrobus2022 <- read_gtfs(file.path(GTFS_path, "2022-04_WMATA.zip"))
MetrobusStops <- stops_as_sf(Metrobus2022$stops) %>%
  st_intersection(., Nova) %>%
  select(stop_id) %>%
  mutate(Agency = "WMATA") %>% unite(stop_id, Agency, col = "newstop_id", remove = F)

#bind with the other nova agency stop data from above
NovaStops_2022wmata <- rbind(NovaStops_2022, MetrobusStops)
```

#### Creating `NovaStopsRoutes` 

To find the amount of service provided, we use the Transit Service Index, or TSI. The formula requires we know the trip frequency at each stop. So using the following handy function, we count up the number of times per *week* a bus or train visits a stop. 


```{r, eval = F}
route_n_stop <- function(gtfszip, agency){
  GTFS_path <- file.path ("Z:",
                          "NVTC General",
                          "Projects and Programs",
                          "Transit Resource Center (TRC)",
                          "Data",
                          "GTFS")
  GTFS <- read_gtfs(file.path(GTFS_path, gtfszip))
  #need service and number of days that service_ids use (multiplier)
  Service <- GTFS$calendar %>%
    mutate(multiplier = monday + tuesday + wednesday + thursday + friday + saturday + sunday)
  #stop_times has both the trip_id and stop_id
  Stops <- GTFS$stop_times %>% select(trip_id, stop_id)
  #routes has trip_id and route_id
  #don't filter by service_id except for VRE
  if (agency == "VRE") {
    Trip <- GTFS$trips %>% filter(service_id == "Regular") %>%
      select(trip_id, route_id, service_id) %>%
      left_join(., Service %>% select(service_id, multiplier))
  } else {
    Trip <- GTFS$trips %>%
      select(trip_id, route_id, service_id) %>%
      left_join(., Service %>% select(service_id, multiplier))
  }
  #what we need is route_id and stop_id, if we join we can bring it all together
  #then select just the columns we need
  Join <- inner_join(Stops, Trip) %>%
    #find count of trips by route_id, stop_id
    group_by(route_id, stop_id, service_id, multiplier) %>% summarize(trips = n()) %>%
    #if service_id is mutli-day, multiply by number of days of service
    mutate(tripfreq = trips*multiplier ) %>%
    #find sum of trip count for each route_id/stop_id
    group_by(route_id, stop_id) %>% summarize(tripfreq = sum(tripfreq)) %>%
    #add column for agency name
    mutate(agency = agency) %>%
    select(route_id, stop_id, tripfreq, agency)
  return(Join)
}
```

Now run that lil function for each agency (minus WMATA, this will be done separately). 

```{r, eval = F}
ARTjoin_2022 <- route_n_stop("2022-04_Arlington.zip", "ART")
CUEjoin_2022 <- route_n_stop("2022-03_CUE.zip", "CUE")
DASHjoin_2022 <- route_n_stop("2022-04_DASH.zip", "DASH")
FFXjoin_2022 <- route_n_stop("2022-03_Fairfax_Connector.zip", "FFX")
LCTjoin_2022 <- route_n_stop("2022-07_Loudoun.zip", "LCT") #ISSUES wTIH TRIPFREQ
VREjoin_2022 <- route_n_stop("2022-03_VRE.zip", "VRE")
```

Now we recreate `route_n_stop` for the slightly more complicated Metrobus: 

```{r, eval = F}
Metrobus2022 <- read_gtfs(file.path(GTFS_path, "2022-04_WMATA.zip"))

#need service and number of days that service_ids use (multiplier)
MetrobusService <- Metrobus2022$calendar %>%
  mutate(multiplier = monday + tuesday + wednesday + thursday + friday + saturday + sunday)

# stops: spatial  filter to nova
MetrobusStops <- stops_as_sf(Metrobus2022$stops) %>%
  #spatially filter to only the stops in va
  st_intersection(.,Nova) %>%
  st_drop_geometry() %>%
  left_join(., Metrobus2022$stop_times) %>% select(trip_id, stop_id)

# trips
MetrobusTrips <- Metrobus2022$trips %>% #filter(service_id == Metrobus_service04) %>%
  select(trip_id, route_id, service_id) %>%
  left_join(., MetrobusService %>% select(service_id, multiplier))

#join stops and trips
MetrobusJoin <- inner_join(MetrobusStops, MetrobusTrips) %>%
  #find count of trips by route_id, stop_id
  group_by(route_id, stop_id, service_id, multiplier) %>% summarize(trips = n()) %>%
  #if service_id is mutli-day, multiply by number of days of service
  mutate(tripfreq = trips*multiplier ) %>%
  #find sum of trip count for each route_id/stop_id
  group_by(route_id, stop_id) %>% summarize(tripfreq = sum(tripfreq)) %>%
  #add column for agency name
  mutate(agency = "WMATA") %>%
  select(route_id, stop_id, tripfreq, agency)
```


...and for Metrorail. But actually we did the route_n_stop analysis outside of R because the GTFS is a bit more complicated. In the next chunk, we just read that data in and reformat it to fit the format of our other `Agencyjoin_2022` variables. The second part of the chunk reads the rail stop data, selects the entrance data, then joins it to the frequency of trip data. 


```{r, eval = F}
#
Metrorailjoin_2022 <- read.csv("AgencyProfileData/Metrorail.csv", sep = ",")%>%
  pivot_longer(cols = 2:4, names_to = "route_id", values_to = "tripfreq") %>%
  rename(station = ï..stop_id) %>% relocate(station, .after = route_id) %>%
  mutate(agency = "WMATA")

#stop locations, using entrance ids,
Metrorailstops <- read.csv(file.path(GTFS_path, "2022-06_WMATA-Railstops.csv")) %>%
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326) %>%
  filter(grepl("ENT_", stop_id)) %>%
  select(ï..VirginiaStation, stop_id) %>% rename(station = ï..VirginiaStation) %>%
  inner_join(., Metrorailjoin_2022) %>% mutate(Mode = "HR", Agency = "WMATA") %>%
  filter(tripfreq != 0) %>%
  unite(station, stop_id, agency, sep = "_", col = "newstop_id") %>%
  unite(route_id, Agency, sep = "_", col = "newroute_id") %>% mutate(AgencyName = "WMATA")
```

Let's bring it all together and create one beautiful `NovaStopsRoutes` file!

```{r, eval = F}
#join all agencies (except rail, this is done after)
Alljoin_2022 <- rbind(ARTjoin_2022,
                      CUEjoin_2022,
                      DASHjoin_2022,
                      FFXjoin_2022,
                      VREjoin_2022,
                      LCTjoin_2022,
                      MetrobusJoin) %>%
  unite(stop_id, agency, col = "newstop_id")

#join with stop data to create spatial dataset (NovaStops data is spatial, Alljoin is not)
#create combined route and stop ids
NovaStopsRoutes_2022 <- merge(Alljoin_2022, NovaStops_2022wmata) %>%
  st_as_sf() %>%
  mutate(Agency = AgencyName) %>%
  unite(route_id, Agency, sep = "_", col = "newroute_id")%>%
  mutate(Mode = ifelse(AgencyName == 'VRE', "CR", "Bus")) %>%
  #merge with rail  stops
  rbind(., Metrorailstops)
```

Okay, we have set up everything we need for both the Access to Transit Service analysis and the Amount of Transit Service analysis. Let's get going! 

## Access to Transit


#### by Jurisidction
Time to do some analysis! We will use `st_interpolate_aw` to find who has access to transit in each jurisdiction. The following function takes all the stops in a route and creates a buffer. We then interpolate the demographic data within that buffered area to find how much access to transit different populations have. 

```{r, eval = T, include = F}
library(knitr)
buffers <- data.frame(
  Mode = c("Bus", "Heavy Rail", "Paratransit", "Commuter Rail"), 
  BufferMiles = c("1/4", "1/2", "3/4", "1"), 
  BufferMeters = c(400, 800, 1200, 1600)
)
```
```{r, eval = T, echo = F}
kable(buffers, caption = "Buffers for Access to Transit")
```


```{r, eval = F}
access2transit <- function(Mode, buff, Census) {

  BuffJoin <- NovaStopsRoutes_2022 %>% filter(Mode == Mode) %>%
    st_buffer(dist = buff) %>% st_union() %>% st_make_valid()

  #interpolate census data to these stops
  TransitCensus <- st_interpolate_aw(
    Census,
    BuffJoin,
    extensive = T
  ) %>% st_drop_geometry() %>% as.data.frame()
  return(TransitCensus)
}
```

We will split up analysis by mode and jurisdiction. But first, let's create polygons for each NVTC jurisdiction. 

```{r, eval = F}
Arl <- countycensus_tract("Arlington") %>% st_transform(crs = 4326) %>% select(-GEOID)
Alx <- countycensus_tract("Alexandria city") %>% st_transform(crs = 4326) %>% select(-GEOID)
CityFFX <- countycensus_tract("Fairfax city") %>% st_transform(crs = 4326) %>% select(-GEOID)
Lou <- countycensus_tract("Loudoun") %>% st_transform(crs = 4326) %>% select(-GEOID)
fc <- countycensus_tract("Falls Church city") %>% st_transform(crs = 4326) %>% select(-GEOID)
ffx <- countycensus_tract("Fairfax County") %>% st_transform(crs = 4326) %>% select(-GEOID)
```

```{r, eval = F}
#arlington
ARlCR <- st_interpolate_aw(
  Arl,
  NovaStopsRoutes_2022 %>% filter(Mode == "CR") %>%
    st_buffer(dist = 1600) %>% st_union() %>% st_make_valid(),
  extensive = T
) %>% mutate(Mode = "CR")
ARlBus <- st_interpolate_aw(
  Arl,
  NovaStopsRoutes_2022 %>% filter(Mode == "Bus") %>%
    st_buffer(dist = 400) %>% st_union() %>% st_make_valid(),
  extensive = T
) %>% mutate(Mode = "Bus")
ARlHR <- st_interpolate_aw(
  Arl,
  NovaStopsRoutes_2022 %>% filter(Mode == "HR") %>%
    st_buffer(dist = 800) %>% st_union() %>% st_make_valid(),
  extensive = T
) %>% mutate(Mode = "HR")
ARlpara <- st_interpolate_aw(
  Arl,
  NovaStopsRoutes_2022 %>%
    st_buffer(dist = 1200) %>% st_union() %>% st_make_valid(),
  extensive = T
) %>% mutate(Mode = "Paratransit")

ArlAccess <- rbind(ARlBus, ARlHR, ARlCR, ARlpara) %>% mutate(Jurisdiction = "Arlington County")
```


Now repeat that same analysis for each jurisdiction. I won't include all the code here to save space, but all you have to do is replace the first input of the `st_interpolate_aw` (in this case, `Arl`) with the relevant county or city. 

Let's combine all those variables and then you're done! 

```{r, eval = F}
NovaAcessJurisdictions <- rbind(LouAccess,
                                ArlAccess,
                                AlxAccess,
                                ffxAccess,
                                CityFFXAccess,
                                fcAccess)
```


In the agency profiles, we show percentages. The denominators of those ratios are created in the next line of code. 

```{r, eval = F}
NovaCensus <- rbind(Arl %>% st_drop_geometry()%>% summarize_all(sum) %>% mutate(County = "Arlington"),
                    Alx %>% st_drop_geometry()%>% summarize_all(sum) %>% mutate(County = "Alexandria"),
                    Lou %>% st_drop_geometry() %>% summarize_all(sum) %>% mutate(County = "Loudoun"),
                    ffx %>% st_drop_geometry()%>% summarize_all(sum) %>% mutate(County = "Fairfax"),
                    fc %>% st_drop_geometry()%>% summarize_all(sum) %>% mutate(County = "Falls Church"),
                    CityFFX %>% st_drop_geometry()%>% summarize_all(sum) %>% mutate(County = "CityFairfax"))

```


## Amount of Service


Now it's time for some amount of service analysis. Instead of looking at who has access to transit, we look at how *much* service each census tract by using the TSI, as mentioned above. Let's get going!

```{r, eval = F}

# Using Census Tracts for NVTC and PRTC jurisdictions
NOVA_Plus <- c("Arlington",
               "Fairfax County",
               "Fairfax city",
               "Loudoun",
               "Alexandria City",
               "Falls Church City",
               "Prince William County",
               "Stafford County",
               "Spotsylvania County",
               "Fredericksburg City",
               "Manassas City",
               "Manassas Park City")

CensusDataTracts <- countycensus_tract(NOVA_Plus) %>%
  st_transform(crs = 4326)

amount_service <- function(newroute_id){
  message(newroute_id)
  #create county shape and transform
  CountyShp <- CensusDataTracts %>% select(GEOID)

  #summarize stop frequency at the route level
  Ag_County <- NovaStopsRoutes_2022 %>% filter(newrt_d == newroute_id) %>%
    group_by(newrt_d, Mode) %>% summarize(stopfreq = mean(tripfrq))

  #buffer the stops, conditional if its a commuter rail (1 mile radius) or bus (1/4 mi)
  Ag_CountyBuff <- Ag_County %>% st_buffer(., ifelse(.$Mode == "CR", 1600,
                                                     ifelse(.$Mode == "HR", 800,400))) %>%
    mutate(area = st_area(geometry))
  #set units to square miles

  Ag_CountyBuff$area <- set_units(Ag_CountyBuff$area, mi^2)
  #calculate area times stop frequency
  #select only numeric values for interpolation
  Ag_CountyBuff <- Ag_CountyBuff %>% mutate(calcTSI = area*stopfreq) %>% select(calcTSI)

  #interpolate "TSI" for each census block group
  CountyInterpol <- st_interpolate_aw(Ag_CountyBuff, CountyShp, extensive = T) %>%
    mutate(area = st_area(.))
  CountyInterpol$area <- set_units(CountyInterpol$area, mi^2)

  #calculate TSI
  CountyInterpol <- CountyInterpol %>% mutate(TSI = calcTSI/area) %>% select(TSI, area)

  #join with census data to have associated GEOIDs
  AmountService <- st_join(CountyInterpol, CountyShp, join = st_equals) %>%
    mutate(newroute_id = newroute_id)
  return(AmountService)
}
```



Since it's route level and we don't want to run like 200 different functions manually, we'll use `lapply` like we did before. 
```{r, eval = F}
routes <- NovaStopsRoutes_2022 %>% distinct(newrt_d, AgncyNm, Mode) %>% as.data.frame()


#run loop through all routes
nova_amountROUTE <- lapply(routes$newrt_d, amount_serviceROUTE) %>%
  do.call(rbind.data.frame, .)
```



## Census 

Load census data. For the transit agency profiles, we are using **tracts**. 
Previous analyses (How the Pandemic Changed Transit) used **block groups**. 


This code varies slightly from the `countycensus` function found in `toolingtransit`. 
As the name indicates, it outputs census tracts instead of block groups. 
```{r, eval = F}
countycensus_tract <- function(County) {
  #load required libraries
  require(tidyverse)
  require(censusapi)
  require(sf)
  require(units)
  #pull census acs data for 2020, select variables for county
  CensusCnty <- get_acs(
    geography = "tract",
    year = 2020,
    variables = c(TotalPopulation = "B01003_001",
                  TotalCommuters = "B08301_001",
                  DriveAlone = "B08301_003",
                  PublicTransport = "B08301_010",
                  HHCount = "B25044_001",
                  PovertyRatioTotal = "C17002_001",
                  PovertyOver200Pct = "C17002_008",
                  HHOwn0Veh = "B25044_003",
                  HHRent0Veh = "B25044_010",
                  WhitePop = "B02001_002"),
    state = "VA",
    county = County,
    geometry = TRUE,
    output = "wide"
  ) %>%
    #calculate new variables
    mutate(Count0Car = HHOwn0VehE + HHRent0VehE,
           NonWhitePop = TotalPopulationE - WhitePopE,
           Under200PctPoverty = PovertyRatioTotalE - PovertyOver200PctE,
           area = st_area(.)) %>%
    #select only necessary variables (remove margin of error columns)
    select(TotalPopulationE, HHCountE, Count0Car, TotalCommutersE, PublicTransportE, NonWhitePop, Under200PctPoverty, area, GEOID)
  #set units for the area as square miles
  CensusCnty$area <- set_units(CensusCnty$area, mi^2)
  return(CensusCnty)
}
```


Now we'll create a  polygon of census tracts for the Northern Virginia region. 
**Note** that this code includes the `GEOID`, which we will need to identify and connect data spatially. 
```{r, eval = F}
NOVA_Counties <-  c("Arlington",
                    "Fairfax County",
                    "Fairfax city",
                    "Loudoun",
                    "Alexandria City",
                    "Falls Church City")
CensusDataTracts <- countycensus_tract(NOVA_Counties)
```


## Service

**Sophie is going to work on adding R code for the Service section, so ignore this section for now!**
